Discrete Quantum Theories and Computing

### 2018-03-02 dissertation.tex
Updates from ShareLaTeX
dissertation.lyx "2.1.1 The single qubit problem" Updating...

\makeatother

@@

A convenience for computing probability and relative state properties
is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\phi}$, that is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\phi}\alpha_{0},\rme^{\rmi\phi}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007},
which we can write down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\phi}$ (``fibering
out by the circle $\Sphere{1}$''). The standard form of this maps
(``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Vec{X}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. 
\begin{figure}[tbh]
\centerline{ \includegraphics[width=0.48\columnwidth]{hopfS2}\hspace{0.03in}
\includegraphics[width=0.48\columnwidth]{hopfFibers}} \centerline{\hspace{1in}
(a) \hfill{}(b) \hspace{1in}} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $\rme^{\rmi\phi}$), each corresponding to
a single point on the sphere in (a). Points in (a) are color coded
corresponding to circles in (b), e.g., one pole contains the red elliptical
circle that would become an infinite-radius circle in a slightly different
projection, and the opposite pole corresponds to the large perfectly
round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}
In Fig.~\ref{hopfFibers.fig}, we illustrate schematically the family
of circles \textit{each one of which is collapsed to a point\/} $(\theta,\phi)$
on the surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{equation}
\left(x_{0},y_{0},x_{1},y_{1}\right)=\left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\ \cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\ \sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right)\,,\label{A8a.eq}
\end{equation}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{equation}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\,.\label{A8b.eq}
\end{equation}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=\rme^{\rmi\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\,,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere,
as shown in Figure \ref{blochSphere.fig}(a).

\begin{figure}[thb]
\begin{centering}
\includegraphics[width=0.47\columnwidth]{blochSphere} \hspace{0.01in}
\includegraphics[width=0.47\columnwidth]{blochSphereArc} 
\par\end{centering}
\centerline{\hspace{1in} (a) \hfill{}(b) \hspace{1in}} \caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Vec{a}$ and $\Vec{b}$ on $\Sphere{2}$,
the shortest rotation carrying the unit normal $\Hat{a}$ to the unit
normal $\Hat{b}$ is the SLERP (spherical linear interpolation) 
\begin{equation}
S(\Hat{a},\Hat{b},t)=\Hat{a}\,\frac{\sin((1-t)\theta)}{\sin\theta}\,+\,\Hat{b}\,\frac{\sin(t\theta)}{\sin\theta}\ ,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\theta$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Figure \ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

@@

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $\rme^{\rmi\phi}$. However,
in practice this is not a construction that is easy to realize in
a practical computation. We now outline an explicit algorithm for
accomplishing the reduction to the irreducible $n$-qubit state space
$\CP{D-1}$; this construction will turn out to be useful for the
validation of our discrete results to follow below.

@@

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\phi}$ from the $n$-qubit states. If we take
a particular weighting of the elements of the density matrix $\rho$,
we can construct a \textit{unit vector\/} of real dimension $D^{2}$
with the form: 
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}

>>>>

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

@@

A convenience for computing probability and relative state properties
is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\frac{\theta}{2}}$, that
is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\frac{\theta}{2}}\alpha_{0},\rme^{\rmi\frac{\theta}{2}}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007},
which we can write down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\frac{\theta}{2}}$
(``fibering out by the circle $\Sphere{1}$''). The standard form
of this maps (``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Vec{X}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $\left(\phi,\psi\right)$ on
the surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

\begin{figure}[tbh]
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.48\columnwidth]{hopfS2}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.48\columnwidth]{hopfFibers}
\par\end{centering}
}\hfill{} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $\rme^{\rmi\frac{\theta}{2}}$), each corresponding
to a single point on the sphere in (a).\protect \\
\todo{{Since }the Hopf fibration{ is fibering out by} the circle
$\Sphere{1}${ }on the three-sphere $\Sphere{3}$ embedded in $\R^{4}$.
Need to confirm whether (b) is a figure in $\R^{4}$ and how to understand
the next sentence.}\protect \\
Points in (a) are color coded corresponding to circles in (b), e.g.,
one pole contains the red elliptical circle that would become an infinite-radius
circle in a slightly different projection, and the opposite pole corresponds
to the large perfectly round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{equation}
\left(x_{0},y_{0},x_{1},y_{1}\right)=\left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\,\sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right)\,,\label{A8a.eq}
\end{equation}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{equation}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\,.\label{A8b.eq}
\end{equation}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=\rme^{\rmi\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\,,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere
called the Bloch sphere, as shown in Fig.~\ref{blochSphere.fig}(a).

\begin{figure}[thb]
\begin{centering}
\hfill{}\subfloat[]{\centering{}\includegraphics[width=0.47\columnwidth]{blochSphere}}\hfill{}\subfloat[]{\begin{centering}
\includegraphics[width=0.47\columnwidth]{blochSphereArc}
\par\end{centering}
}\hfill{}
\par\end{centering}
\caption{(a) The conventional Bloch sphere with a unique state represented
by the point at the red sphere. (b) The geodesic shortest-distance
arc connecting two one-qubit quantum states.}
\label{blochSphere.fig} 
\end{figure}

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Vec{a}$ and $\Vec{b}$ on $\Sphere{2}$,
the shortest rotation carrying the unit normal $\Hat{a}$ to the unit
normal $\Hat{b}$ is the SLERP (spherical linear interpolation) 
\begin{equation}
S(\Hat{a},\Hat{b},t)=\Hat{a}\,\frac{\sin((1-t)\theta)}{\sin\theta}\,+\,\Hat{b}\,\frac{\sin(t\theta)}{\sin\theta}\ ,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\theta$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Fig.~\ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

@@

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $\rme^{\rmi\frac{\theta}{2}}$.
However, in practice this is not a construction that is easy to realize
in a practical computation. We now outline an explicit algorithm for
accomplishing the reduction to the irreducible $n$-qubit state space
$\CP{D-1}$; this construction will turn out to be useful for the
validation of our discrete results to follow below.

@@

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\frac{\theta}{2}}$ from the $n$-qubit states.
If we take a particular weighting of the elements of the density matrix
$\rho$, we can construct a \textit{unit vector\/} of real dimension
$D^{2}$ with the form: 
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}
